Combinig geological and ecological proxies to model latitudinal temperature gradients in the early Eocene greenhouse

### 2022-12-02 manuscript/manuscript.Rmd
introduction writing

# Introduction
The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021]. Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018]. Reconstructions of palaeoclimate commonly follow one of two approaches. Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g. @Liu2009]. To generate regional or global average estimates, the mean of multiple local estimates is taken [e.g. @Veizer2015]. This approach has the advantage of being transparent and easy to reproduce, modify or expand with new data, but is prone to sampling bias [@Jones2022]. Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???]. However, this approach still relies on proxy data to constrain the large number of possible model setups and ground truth the model output [@???] It also requires substantial computing resources and technical expert knowledge, and is thus difficult to implement or modify for most palaeoclimate researchers.

Climate reconstructions of low - intermediate complexity provide an alternative that can be employed more readily by a broad range of users. Examples of this approach include proxy compilations that take into account the spatial nature of the data, e.g. @Hollis2019 investigated latitudinal temperature trends across the late Palaeocene - early Eocene with a spatially resolved proxy record. Expanding from this approach, some proxy-based reconstructions use statistical modelling to infer climatic patterns, and to bridge data gaps, e.g. using polynomials to reconstruct latitudinal gradients [@Bijl2009], or Gaussian process regression for a 2D-reconstruction of surface temperatures [@Inglis2020]. These latter approaches take advantage of the autoregressive nature

Hook -> Generic problem:
- Reconstructing past climate important for understanding current/future global warming
- Two main pathways: highly complex ESM <--> proxy compilations
- Problems with proxy data: only provides a local picture (few local sites interpreted as global signal)
- Problems with ESM (long time to run, very complex, a lot of assumptions on processes/doesn't consider how they may change, need to be constrained with proxy data)
- Lack of intermediate-complexity models to make full use of the proxy data

Specific problem:
- exemplified with the EECO: proxies and ESM hard to reconcile, EECO important climate analogue
- a lot of proxies recrystallised <--> unreliable data or lack of data

@@

- EECO specific information
- Disagreement between GCM + geochem data

@@

-->
# Materials & Methods

@@

[Fig. 3]
- Modern data
- Sample based on Eocene localities

@@

## Processing of model results
- Calculate lat gradient
- Calculate global T
[Fig. 4]

Modelled sea surface temperature estimates were generated with eq. ?, calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior. The median, $2.5^{th}$ and $97.5^{th}$ percentile of temperatures where then taken from all temperature estimates obtained at the latitude of interest.

@@

## Model validation
- Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
- Show that the modern gradient can be reproduced with Eocene sampling distribution
[Fig 3 - Modern gradient]
The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3a). Degrading the amount and distribution of data points to match the early Eocene sampling distribution still results in a close match of the modelled gradient with the empirical gradient (Fig 3b). A noticable, but minor offset is apparent in high latitudes, where sampling is most incomplete (Fig 3b).

## EECO reconstruction
- Show Eocene gradient
- Global t : use raster package for calculation of areal weights
- report lat gradient

```{r results block - for in-line code, echo = FALSE}
setwd('..') # move up to parent folder of /manuscript

@@

## Fig 4 - compare Eocene gradient with previous reconstructions / ES models

>>>>

# Introduction
The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021]. Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018]. Reconstructions of palaeoclimate commonly use 1) geochemical proxy data, 2) ecological or lithological climate indicators, or 3) earth system modelling.

Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g. @Liu2009]. To generate regional or global average estimates, the mean of multiple local estimates is taken [e.g. @Veizer2015]. This approach has the advantage of being transparent and easy to reproduce, modify or expand with new data, but is prone to sampling bias [@Jones2022]. Ecological and lithological climate data provide an alternative method of reconstructing palaeotemperatures, using e.g. leaf physiognomy and floral assemblages to derive palaeotemperature proxies [@Greenwood2003], or using lithological indicators to determine the range of palaeoclimatic belts [@Scotese2021]. Despite some recent advances [@Chandra2021], ecological and lithological climate indicators are still rarely used in quantitative palaeoclimate reconstructions, particularly for the marine realm. Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???]. However, this approach still relies on proxy data to constrain the large number of possible model setups and ground truth the model output [@???] It also requires substantial computing resources and technical expert knowledge, and is thus difficult to implement or modify for most palaeoclimate researchers.

Spatially resolved climate reconstructions provide an alternative that can be employed more readily by a broad range of users. Examples of this approach include the visualisation of proxy compilations along a latitudinal gradient [@Hollis2019, @Vickers2021]. Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].

Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.

Hook -> Generic problem:
- Reconstructing past climate important for understanding current/future global warming
- Two main pathways: highly complex ESM <-> proxy compilations
- Problems with proxy data: only provides a local picture (few local sites interpreted as global signal)
- Problems with ESM (long time to run, very complex, a lot of assumptions on processes/doesn't consider how they may change, need to be constrained with proxy data)
- Lack of intermediate-complexity models to make full use of the proxy data

Specific problem:
- exemplified with the EECO: proxies and ESM hard to reconcile, EECO important climate analogue
- a lot of proxies recrystallised <-> unreliable data or lack of data

@@

- EECO specific information
- Disagreement between GCM + geochem data
-->

@@

Our hierarchical approach accounts for different sample sizes between and uncertainties within proxy sites, and

-->

# Materials & Methods

@@

[Fig. 3]
<!--
- Modern data
- Sample based on Eocene localities
-->

@@

## Processing of model results
<!--
- Calculate lat gradient
- Calculate global T
[Fig. 4]
-->
Modelled sea surface temperature estimates were generated with eq. ?, calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior. The median, $2.5^{th}$ and $97.5^{th}$ percentile of temperatures where then taken from all temperature estimates obtained at the latitude of interest.

@@

## Model validation
<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
- Show that the modern gradient can be reproduced with Eocene sampling distribution
-->
[Fig 3 - Modern gradient]
The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3a). Degrading the amount and distribution of data points to match the early Eocene sampling distribution still results in a close match of the modelled gradient with the empirical gradient (Fig 3b). A noticable, but minor offset is apparent in high latitudes, where sampling is most incomplete (Fig 3b).

## EECO reconstruction
<!-- - Show Eocene gradient
- Global t : use raster package for calculation of areal weights
- report lat gradient
-->
```{r results block - for in-line code, echo = FALSE}
setwd('..') # move up to parent folder of /manuscript

@@

## Fig 4 - compare Eocene gradient with modern gradient, also with previous reconstructions / ES models?
